High-order cognition is supported by complex but compressible brain activity patterns

### 2023-02-23 paper/main.tex

To the extent that brain activity patterns contain rich task-relevant
information, we should be able to use the activity patterns to accurately
differentiate between different aspects of the task~\citep[e.g., using pattern
classifiers;][]{NormEtal06b}. For example, prior work has shown a direct
correspondence between classification accuracy and the information content of a
signal~\citep{Alva02}. To the extent that brain activity patterns are
compressible, we should be able to generate simplified (e.g., lower
dimensional) representations of the data while still preserving the relevant or
important aspects of the original signal. In general, information content and
compressibility are related but are partially dissociable
(Fig.~\ref{fig:information--compression}).

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{figs/pie_example.pdf}
  \caption{\textbf{Illustration of compression.} Visual analogy for
    neural compression. Here are 2 images of pies, one more complex
    than the other. \textbf{a.} It takes fewer components to reach the same percent
    variance explained in the less complex pie, which corresponds to
    higher compression. \textbf{b.} However, with very few components, similar variance is explained in both pies.
\textbf{c.} Plots the cumulative explained variance for more and more components.}
  \label{fig:information-compression}
\end{figure}

Specifically, if a given signal
(e.g., a representation of brain activity patterns) contains more information
about ongoing cognitive processes, then the peak decoding accuracy should be
high. And if the signal is compressible, a low-dimensional embedding of the
signal will be similarly informative to the original signal. A signal can be
informative but not compressible, informative but

%   - e.g., you could have something that is highly compressible (just a few components are sufficient to achieve some threshold information level) but
%     *also* very rich-- e.g., the maximum number of informative components is large
% compression and task demands -- MackEtal20, ZhouEtal19
% superEEG -- prediction of held-out data; implies redundancy
% timecorr -- higher order network patterns during higher order cognition implies more complexity
% cognitive richness and scrambling-- various Hasson lab papers

We're interested in the complexity of brain patterns that underly different
types of thoughts. To explore this question space, we will take brain patterns
recorded under different experimental conditions used in Aim 2, and project
them into lower dimensional spaces using principle components analysis. We can
then ask how well those low-dimensional embeddings of the data retain
cognitively relevant information like when in a story someone is listening to.

This work has been inspired, in part, by~\cite{MackEtal20}. In this paper, they
investigated the role of the prefrontal cortex in filtering out irrelevant
content. Specifically, they looked at if the vmPFC performs data reduction on
incoming information through compression. This was motivated, in part, by
orbital frontal cortex (OFC) compression in rats~\citep{ZhouEtal19}. They
studied this using a learning paradigm in which participants had to classify
insects based on different numbers of feature dimensions. The idea was that
participants in some learning blocks, participants could identify the insects
based on one feature (low complexity) or several features (high complexity),
but importantly the stimuli remained the same across all learning problems.
They found that complexity and compression had an inverse relationship; the
lower complexity of a conceptual space, the higher the degree of compression.
Building on this idea, we wonder if varying degrees of compression is performed
throughout the brain. We also want to test this idea, but using varying levels
of engagement listening to a naturalistic stimuli.

To understand the degree of compression throughout the brain during cognition,
we will use the same fMRI data from Aim 2, collected while participants
listened to a story in different scrambling conditions. We will measure the
degree that multivoxel activation patterns are compressed during story
listening using principle components analysis (PCA) a method for low-rank
approximation of multidimensional data~\citep{EckaYoun36}. We will explore this
using decoding accuracy as a function of the number of components, or
dimensions, in the low-dimensional space under different cognitive conditions.

You can imagine two reasonable predictions of how cognition is reflected in
brain patterns. The first is as our thoughts become more complex, they are
supported by more complex brain patterns, and require more components to
decode. The second is that when thoughts are deeper and more complicated, the
units of neural activity would carry more information, and would require
therefore fewer components to decode.

This idea can be explored in this visual analogy (Fig.~\ref{fig: pie_example})
for neural compression. Here there are two images of pies, the top pie is more
complex than the bottom. On the left we're illustrating that it takes fewer
components to reach the same 95 percent variance explained in the less complex
pie, which corresponds to higher compression. However, on the right with very
few components similar variance is explaining both pies.

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{figs/pie_example.pdf}
  \caption{\textbf{Illustration of compression.} Visual analogy for
    neural compression. Here are 2 images of pies, one more complex
    than the other. \textbf{a.} It takes fewer components to reach the same percent
    variance explained in the less complex pie, which corresponds to
    higher compression. \textbf{b.} However, with very few components, similar variance is explained in both pies.
\textbf{c.} Plots the cumulative explained variance for more and more components.}
  \label{fig:pie_example}
\end{figure}

We investigated the dimensionality of neural patterns by training
classifiers using more and more principle components. Or, in other
words, we used less and less compression to decode.
We applied the approach to
a neuroimaging dataset comprising data collected as participants
listened to a story varying in cognitive richness~\citep{SimoEtal16}.

% PCA, or the closely related factor analysis and singular value decomposition (SVD) (Hastie et al., 2009), is widely used in the study of individual differences and aids estimating how many latent components, or “factors”, underlie a pattern of (item) responses within or across participants, as for instance in the context of intelligence (Spearman, 1904) or personality tests (Cattell, 1947). In the context of neuroimaging, PCA has been used to identify brain networks (Huth et al., 2012; Friston et al., 1993).

 % add citation:
% S.M.Smith,A.Hyva ̈rinen,G.Varoquaux,K.L.Miller,and C. F. Beckmann, “Group-PCA for very large fMRI datasets,” NeuroImage, vol. 101, pp. 738, 2014.

\subsection*{Evaluation metrics}
 We will evaluate the degree of
compression of held-out
neuroimaging data by assessing the time at which it was collected.  We will use this evaluation (timepoint
decoding) as a proxy for
gauging how much explanatory power the compressed data held
with respect to the observed data.

\subsubsection*{Timepoint decoding}

To explore how compression varies with complexity, we will use a previous neuroimaging dataset
~\cite{SimoEtal16}  in which participants listened to an audio recording of a story; 36 particpants listen to an intact version of the story, 17 participants listen to time-scrambled recordings of the same story where paragraphs were scrambled, 36 particpants listen to word-scrambled version and 36 participants lay in rest condition.

Following the analyses conducted by (HTFA)~\cite{MannEtal18}, we first
apply \textit{hierarchical topographic factor analysis} (HTFA) to
the fMRI datasets to obtain a time series of 700 node activities for
every participant.  We then apply dimensionality reduction
(Incremental PCA) for each group.

We then compare the groups’ activity patterns (using Pearson
correlations) to estimate the story times each corresponding pattern
using more and more principle components.

To assess decoding accuracy, we randomly divide participants for each
stimulus into training and testing groups. We then compare the
groups’ activity patterns (using Pearson correlations) to estimate the
story times each corresponding pattern using more and more principle
components (as the data became less compressed). Specifically, we ask, for each timepoint: what are the correlations
between the first group's and second group's activity patterns at each
order. We note that the decoding test we used is a conservative in which we count a timepoint label as incorrect if it is not an exact match.

>>>>

To the extent that brain activity patterns contain rich task-relevant
information, we should be able to use the activity patterns to accurately
differentiate between different aspects of the task~\citep[e.g., using pattern
classifiers;][]{NormEtal06b}. For example, prior work has shown a direct
correspondence between classification accuracy and the information content of a
signal~\citep{Alva02}. To the extent that brain activity patterns are
compressible, we should be able to generate simplified (e.g., lower
dimensional) representations of the data while still preserving the relevant or
important aspects of the original signal. In general, information content and
compressibility are related but are partially dissociable
(Fig.~\ref{fig:information-compression}).

\begin{figure}[tp]

\centering
\includegraphics[width=\textwidth]{figs/information_and_compressibility}

\caption{\textbf{Information content and compressibility.} \textbf{A. Variance
explained for two images.} We applied principal components analysis to a
photograph and drawing, treating each row of the images as ``observations.''
Across columns, we identified the number of components required to explain
100\%, 95\%, 75\%, or 50\% of the cumulative variance in each image (the 100\%
columns denote the original images). The numbers of components are indicate in
parentheses, and the resulting ``compressed'' images are displayed. \textbf{B.
Representing two images with different numbers of components.} Using the same
principal component decompositions as in Panel A, we computed the cumulative
proportion of variance explained with 380 (original images), 20, 4, or 2
components. \textbf{C. Cumulative variance explained versus number of
components.} For the images displayed in Panels A and B, we plot the cumulative
proportion of variance explained as a function of the number of components used
to represent each image. \textbf{D. Information rate and compressibility.}
Across multiple images, the information rate (i.e., the amount of information
contained in each image; horizontal axis) is high when each individual pixel
provides information that cannot be inferred from other pixels.
High-information rate images tend to be high-resolution, and low-information
rate images tend to be low-resolution. Compressibility is related to the
difference between the information required to specify the original versus
compressed images (vertical axis). Highly compressible images often contain
predictable structure (redundancies) that can be leveraged to represent the
images much more efficiently than in their original feature spaces.}

\label{fig:information-compression} 
\end{figure}

% Specifically, if a given signal
% (e.g., a representation of brain activity patterns) contains more information
% about ongoing cognitive processes, then the peak decoding accuracy should be
% high. And if the signal is compressible, a low-dimensional embedding of the
% signal will be similarly informative to the original signal. A signal can be
% informative but not compressible, informative but

% %   - e.g., you could have something that is highly compressible (just a few components are sufficient to achieve some threshold information level) but
% %     *also* very rich-- e.g., the maximum number of informative components is large
% % compression and task demands -- MackEtal20, ZhouEtal19
% % superEEG -- prediction of held-out data; implies redundancy
% % timecorr -- higher order network patterns during higher order cognition implies more complexity
% % cognitive richness and scrambling-- various Hasson lab papers

% We're interested in the complexity of brain patterns that underly different
% types of thoughts. To explore this question space, we will take brain patterns
% recorded under different experimental conditions used in Aim 2, and project
% them into lower dimensional spaces using principle components analysis. We can
% then ask how well those low-dimensional embeddings of the data retain
% cognitively relevant information like when in a story someone is listening to.

% This work has been inspired, in part, by~\cite{MackEtal20}. In this paper, they
% investigated the role of the prefrontal cortex in filtering out irrelevant
% content. Specifically, they looked at if the vmPFC performs data reduction on
% incoming information through compression. This was motivated, in part, by
% orbital frontal cortex (OFC) compression in rats~\citep{ZhouEtal19}. They
% studied this using a learning paradigm in which participants had to classify
% insects based on different numbers of feature dimensions. The idea was that
% participants in some learning blocks, participants could identify the insects
% based on one feature (low complexity) or several features (high complexity),
% but importantly the stimuli remained the same across all learning problems.
% They found that complexity and compression had an inverse relationship; the
% lower complexity of a conceptual space, the higher the degree of compression.
% Building on this idea, we wonder if varying degrees of compression is performed
% throughout the brain. We also want to test this idea, but using varying levels
% of engagement listening to a naturalistic stimuli.

% To understand the degree of compression throughout the brain during cognition,
% we will use the same fMRI data from Aim 2, collected while participants
% listened to a story in different scrambling conditions. We will measure the
% degree that multivoxel activation patterns are compressed during story
% listening using principle components analysis (PCA) a method for low-rank
% approximation of multidimensional data~\citep{EckaYoun36}. We will explore this
% using decoding accuracy as a function of the number of components, or
% dimensions, in the low-dimensional space under different cognitive conditions.

% You can imagine two reasonable predictions of how cognition is reflected in
% brain patterns. The first is as our thoughts become more complex, they are
% supported by more complex brain patterns, and require more components to
% decode. The second is that when thoughts are deeper and more complicated, the
% units of neural activity would carry more information, and would require
% therefore fewer components to decode.

% This idea can be explored in this visual analogy (Fig.~\ref{fig: pie_example})
% for neural compression. Here there are two images of pies, the top pie is more
% complex than the bottom. On the left we're illustrating that it takes fewer
% components to reach the same 95 percent variance explained in the less complex
% pie, which corresponds to higher compression. However, on the right with very
% few components similar variance is explaining both pies.

% We investigated the dimensionality of neural patterns by training
% classifiers using more and more principle components. Or, in other
% words, we used less and less compression to decode.
% We applied the approach to
% a neuroimaging dataset comprising data collected as participants
% listened to a story varying in cognitive richness~\citep{SimoEtal16}.

% % PCA, or the closely related factor analysis and singular value decomposition (SVD) (Hastie et al., 2009), is widely used in the study of individual differences and aids estimating how many latent components, or “factors”, underlie a pattern of (item) responses within or across participants, as for instance in the context of intelligence (Spearman, 1904) or personality tests (Cattell, 1947). In the context of neuroimaging, PCA has been used to identify brain networks (Huth et al., 2012; Friston et al., 1993).

%  % add citation:
% % S.M.Smith,A.Hyva ̈rinen,G.Varoquaux,K.L.Miller,and C. F. Beckmann, “Group-PCA for very large fMRI datasets,” NeuroImage, vol. 101, pp. 738, 2014.

% \subsection*{Evaluation metrics}
%  We will evaluate the degree of
% compression of held-out
% neuroimaging data by assessing the time at which it was collected.  We will use this evaluation (timepoint
% decoding) as a proxy for
% gauging how much explanatory power the compressed data held
% with respect to the observed data.

% To explore how compression varies with complexity, we will use a previous neuroimaging dataset
% ~\cite{SimoEtal16}  in which participants listened to an audio recording of a story; 36 particpants listen to an intact version of the story, 17 participants listen to time-scrambled recordings of the same story where paragraphs were scrambled, 36 particpants listen to word-scrambled version and 36 participants lay in rest condition.

% Following the analyses conducted by (HTFA)~\cite{MannEtal18}, we first
% apply \textit{hierarchical topographic factor analysis} (HTFA) to
% the fMRI datasets to obtain a time series of 700 node activities for
% every participant.  We then apply dimensionality reduction
% (Incremental PCA) for each group.

% We then compare the groups’ activity patterns (using Pearson
% correlations) to estimate the story times each corresponding pattern
% using more and more principle components.

% To assess decoding accuracy, we randomly divide participants for each
% stimulus into training and testing groups. We then compare the
% groups’ activity patterns (using Pearson correlations) to estimate the
% story times each corresponding pattern using more and more principle
% components (as the data became less compressed). Specifically, we ask, for each timepoint: what are the correlations
% between the first group's and second group's activity patterns at each
% order. We note that the decoding test we used is a conservative in which we count a timepoint label as incorrect if it is not an exact match.
